Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model

### 2021-10-08 analysis/paper/first_draft_rev-1/paper.Rmd
incorporate AGH suggestions

---
title: "Investigating Biases Associated with Dietary Starch Incorporation into an Oral Biofilm Model"
author:
  - Bjørn Peare Bartholdy
  - Amanda G. Henry
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    #template: "../templates/sn-article-template/sn-article.tex"
    includes:
      in_header: "../templates/latex/pre-print/preamble.tex"
      before_body: "../templates/latex/pre-print/before_body.tex"
    df_print: kable
    toc: no
    latex_engine: xelatex
    fig_caption: yes
    number_sections: yes
  bookdown::html_document2:
    df_print: kable
    keep_md: true
  bookdown::word_document2:
    reference_docx: "../templates/template.docx"
linestretch: 2.0 # double-spacing for review purposes
bibliography: refs.bib
csl: apa.csl
abstract: |
  Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
keywords: 
  - oral biofilm
  - starch
  - alpha-amylase
  - dental calculus
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.path = "../figures/")
library(tidyverse)
library(broom)
library(here)
library(patchwork)
# load R scripts
source(here("analysis/scripts/analysis.R"))
source(here("analysis/scripts/tables.R"))
source(here("analysis/scripts/figures.R"))
source(here("analysis/scripts/amylase-assay.R"))
## Compile refs.bib using rbbt package ##
cite_keys <- rbbt::bbt_detect_citations(here("analysis/paper/paper.Rmd")) 
#cite_keys <- unlist(cite_keys, use.names = FALSE)
rbbt::bbt_write_bib(here("analysis/paper/refs.bib"), keys = cite_keys, overwrite = T)
## end refs.bib ##
```

@@

<!-- more detailed background, comprehensible to scientists in related disciplines -->
Until recently, only a few studies directly investigated the presence of plant
microremains
in the dental calculus of archaeological remains. The ability to extract phytoliths
from the dental calculus of archaeological fauna to investigate diet was first 
noted by Armitage [-@armitageExtractionIdentificationOpal1975], 
and later by Middleton and Rovner [-@middletonExtractionOpalPhytoliths1994],
and Fox and colleagues [-@foxPhytolithCalculus1996]. Starches and phytoliths were
extracted from human dental calculus by Cummings and Magennis [-@cummingsMayanCalculus1997].  
In more recent years, the study of dental calculus has increased exponentially, 
and the wealth of information contained within the mineralised matrix has been, 
to a larger extent, acknowledged. The applications of dental calculus span a wide
variety of archaeological research areas, such as oral microbiome 
characterisation (including pathogens) through the analysis of DNA and proteins 
[@adlerSequencingAncientCalcified2013; @warinnerPathogensHostImmunity2014],
microbotanical remains [@henryCalculusSyria2008; @hardyStarchGranulesDental2009;
@mickleburghNewInsightsConsumption2012], 
other organic residues and proteins from dietary compounds 
[@buckleyDentalCalculusCooking2014; @hendyProteomicCalculus2018], 
and nicotine-use [@eerkensDentalCalculusSource2018]. Especially the extraction 
of starch granules has become a rich source of dietary
information, as starch granules have proven to preserve well within dental calculus
over a variety of geographical and temporal ranges 
[@pipernoStarchGrains2008; @henryNeanderthalCalculus2014; @taoWheatCalculus2020;
@jovanovicNeolithicCalculus2021].

@@

```{r protocol-fig, fig.cap="Overview of experiment protocol"}
knitr::include_graphics(here("analysis/figures", "ExpL_protocol.png"))
```

<!-- modify sampling days from protocol picture to match amylase sampling -->

@@

An $\alpha$-amylase (EC 3.2.1.1) activity assay was conducted on artificial
saliva samples collected from the plate wells on days 3, 5, 6, 9, and 12. 
Whole saliva samples were collected on days 0, 3, and 5 as positive controls.
Collected samples were stored at 4&deg;C until the assay was conducted on day 18. 
All samples and standard curves were run in triplicates on two separate plates.
Positive control saliva samples were compared against a standard curve containing 
H~2~O, while artificial saliva samples were compared against a standard curve
containing sterile artificial saliva (due to the colour of artificial saliva).
Two photometric readings were conducted for each plate with a 540 nm filter on a 
Multiskan FC Microplate Photometer (Thermo Scientific 51119000). 
The protocol is a slightly modified version of an Enzymatic Assay of $\alpha$-Amylase
(https://www.sigmaaldrich.com/NL/en/technical-documents/protocol/protein-biology/enzyme-activity-assays/enzymatic-assay-of-a-amylase) [@bernfeldAmylase1955], which measures the amount of 
maltose released from starch by $\alpha$-amylase activity. Results are reported
in units (U) per mL enzyme, where 1 U releases 1 mg of maltose.
For a more detailed protocol, see <!-- protocols.io DOI -->

@@

Extraction of starches from the calculus samples was performed by dissolving the
calculus in -0.5 M<!-- what percentage???--> ethylenediaminetetraacetic acid (EDTA) 
[@trompEDTACalculus2017; @lemoyneCalculusPretreatments2021; @modiCalculusMethodologies2020], 
and vortexing for 3 days until the sample was completely dissolved. 
Twenty $\mu$l of sample was mounted onto a slide with an 18x18 mm coverslip.
When transferring the sample to the slide, the sample was homogenised using 
the pipette to ensure that the counted transects were representative of the 
whole slide. Wheat and mix counts were extrapolated as described above.
Both wheat and potato granules were divided into three size categories: 
small (<10 $\mu$m), medium (10 -- 20 $\mu$m), and large (>20 $\mu$m).

@@

All samples yielded sufficient biofilm growth and starch incorporation to be 
included in the analysis, resulting in a total of 48 biofilm samples (two plates of 24),
45 of which were used for analysis (three samples were set aside for later
 analysis).
Most control samples contained no starch granules, while some contained negligible
quantities (see Supplementary Material).

@@

```{r amylase-activity}
# table of results reported in units amylase per mL enzyme (but let's be honest,
  # ...it doesn't really matter what the unit is. No activity is no activity)
cols <- c("1", "2", "3")  
rows <- c("S1", "S2", "S3", "B1", "B2", "B3", "B4", "B5", "BT1", "BT2", "BT3")
plt1_ph1_result <- rbind(sal1_ph1, bmm1_ph1)
  plt1_ph1_result[plt1_ph1_result < 0] <- 0
```

```{r amylase-activity-tbl}
rownames(plt1_ph1_result) <- rows
colnames(plt1_ph1_result) <- cols
knitr::kable(signif(plt1_ph1_result, 3), row.names = TRUE, caption = "Amylase activity in U/mL enzyme, where a U is mg maltose released from starch in six minutes at 36 degrees Celsius. Plate 1, photometric reading 1. Negative values converted to 0.")
```

No appreciable amount of $\alpha$-amylase activity was detected in any of the artificial
saliva samples from any of the days that were sampled. Only positive controls
contained amylase activity that could be detected in the assay (Table
\@ref(tab:amylase-activity-tbl)). The results from the additional plates and
photometric readings can be found in the 
[Supplementary Materials](../supplementary-materials/supp-mat.Rmd). The results
are not comparable to other studies presenting $\alpha$-amylase activity levels
in humans; however, they are sufficient to show that there is no activity in the
system.

@@

<!-- separated treatments -->

@@

```{r sep-plot, fig.cap="(A) Proportion (%) of sizes of starch granules in the wheat solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and (B) in the potato solution (outer ring) and extracted from the potato-treatment samples (inner ring)."}
#plot_grid(pl_wheat1, pl_potato1, labels = c("A", "B"), label_y = 0.1)
pl_wheat1 + pl_potato1 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

```{r mix-plot, fig.cap="Proportion of sizes of (A) wheat granules in the mixed solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and sizes of (B) potato granules in the solution (outer ring) and extracted from the potato-treatment samples (inner ring)."}
#plot_grid(pl_wheat2, pl_potato2, labels = c("C", "D"), label_y = 0.1)
pl_wheat2 + pl_potato2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

<!-- Combine to a single figure? -->

@@

<!-- separated and mixed treatments -->
The difference in incorporation between the size categories resulted in a change
in size ratios between the original starch solutions and the extracted samples.
Large potato granules (> 20 $\mu$m) were most affected, with a 
`r paste0(abs(diff_sep$potato[3]), "%")` 
decrease in relative abundance in the potato-only treatment, and a
`r paste0(abs(diff_mix$potato[3]), "%")` 
decrease in mixed treatments. Medium granules increased in relative abundance 
across all samples, while small granules decreased in wheat treatments and 
increased in potato treatments 
(Figures \@ref(fig:sep-plot) and \@ref(fig:mix-plot)).

<!-- Chi square test for size counts? -->

@@

<!--
Given the low proportion of starch granule retention in our dental calculus model,
and the limited difference between treatments, we can conjecture that starch 
availability is not a main factor in starch incorporation. While there was a
proportional incorporation, i.e. more granules in the solution led to more granules
in the calculus, the difference between solutions [a number/percentage] was of a
higher magnitude than the difference between samples <!-- a number/percentage-->

<!-- Scientific findings as extension of previous work -->
Previous research conducted on dental calculus from contemporary humans and non-human
primates suggest a high level of stochasticity involved in the retention of 
starch granules in dental calculus, and that starch granules extracted from dental 
calculus are underrepresented with regard
to actual starch intake, which is consistent with our findings (illustrated by high
standard deviations and low proportional incorporation).
Leonard and colleagues [-@leonardDentalCalculus2015] found individual
calculus samples to be a poor predictor of diet in a population, as many of the 
consumed plants were missing from some individual samples, but were present in others.   
Power and colleagues [-@powerChimpCalculus2015] 
presented similar findings in non-human primates, where phytoliths were more
representative of individual diets than starch granules.

<!-- secondary findings -->
We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 $\mu$m in 
diameter were underrepresented in the calculus samples compared to the original 
starch solutions, an effect that was consistent across all three treatments, 
while medium granules (10 -- 20 $\mu$m) were often over-represented
(Table \@ref(tab:sample-prop-tbl), 
and 
Figures \@ref(fig:sep-plot) and \@ref(fig:mix-plot)).
This was especially true for the potato starches, which can reach up to 100 $\mu$m
in diameter, whereas wheat generally only reach up to 35 $\mu$m 
[@gismondiStarchGranulesData2019; @haslamDecompositionStarch2004].
The size bias is consistent with the findings by Power and colleagues
[-@powerChimpCalculus2015],
which found that plants producing starches 10--20 $\mu$m in
size were over-represented; although, the representation of granules larger than
20 $\mu$m is unclear.  
Given the differences in proportional incorporation between wheat and potato,
especially for small and medium granules, it is also possible that the morphology
of granules plays a role. However, the proportional count differences between large 
wheat and potato granules were less pronounced.

<!-- Since those results were obtained -->
<!-- on a modern population of chimps that were only buried for a short period of time,  -->
<!-- degradation was likely limited [@powerChimpCalculus2015]. -->

The taphonomy of starch granules has been addressed previously, and depends on 
burial environment (pH, temperature, moisture content, microorganisms), as well 
as processing of the granules prior to deposition 
[@henryCookingStarch2009; @graneroStarchTaphonomy2020; @haslamDecompositionStarch2004]. 
The potential for detrimental diagenetic effects on starch granules increases as 
the size and amylose content decreases
[@francoStarchDegradation1992; @haslamDecompositionStarch2004].

<!-- add potential for morphological differences affecting incorporation 
 Medium potato starches had the highest rate of incorporation 
 All large starches had a consistently low rate of incorporation 
 Starch granule adhesion to bacteria and enzymes? Do they have a preference?
 Binding to select species in a biofilm could explain clustering
 Colonisation of bacteria onto starch @nagaraBacteriaNutrients2017-->

@@

Another potentially important factor 
from our results is the size of the calculus deposit, which seems to influence the 
quantity of starches extracted from the calculus, as we found a strong positive 
correlation between biofilm size and retained starch granules (Figure \@ref(fig:cor-plot));
a result that contradicts findings from archaeological contexts
[@wesolowskiEvaluatingMicrofossilContent2010; @dudgeonDietGeographyDrinking2014]. 
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \@ref(fig:cor-plot2));
while the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches. 
The lack of a correlation
in archaeological contexts could be due to diagenetic effects. Alternately, the absence of 
amylase activity in our study could be impacting our interpretations.

<!-- How are starch granules trapped in dental calculus? Does anyone actually know? - adhesion to bacteria
- stuck in cracks and crevices
- covered by new layer of plaque

-->
The mechanism by which starch granules are incorporated into plaque and calculus 
remains largely unknown, and few studies have directly investigated potential
mechanisms. We know that a proportion of the starch granules entering 
the mouth can become trapped in the plaque/calculus, and can be recovered from
archaeological samples of significant age 
[@henryNeanderthalCalculus2014; @buckleyDentalCalculusCooking2014; @wuDietEarliest2021]. 
Studies have also shown that not all starch granules come from a dietary source. 
Other pathways include cross-contamination from plant interactions in soil, such 
as palm phytoliths adhering to the skin of sweet potatoes 
[@trompDietaryNondietary2015], 
or accidental ingestion not related to food consumption 
[@radiniFoodMultiplePathways2017; @radiniMedievalWomenEarly2019].  
When starch granules enter the mouth, whether through ingestion of food or accidental
intake, they immediately encounter multiple obstacles. It is likely
that the bulk of starch granules are swallowed along with the food, and are 
only briefly present in the oral cavity. Other granules that are broken off
during mastication may be retained in the dentition.

<!-- initial incorporation of starch granules? 
cracks and crevices? attachment to bacteria? -->

These granules are then
susceptible to mechanical removal by the tongue, salivary clearance, and hydrolysis
by $\alpha$-amylase [@kashketFoodParticles1996].
Starch granules that are trapped in crevices and channels on the surface of the
mineralised plaque are (at least to some extent) protected from salivary
clearance and mechanical cleaning actions of the tongue and lips, especially once
a new layer of plaque has covered the surface of the calculus. 
The size bias against large granules (>20 $\mu$m) from both wheat and potato 
(Table \@ref(tab:sample-prop-tbl)) may give further credence to
this incorporation pathway, as the smaller starch granules have an advantage over
larger granules, and can be stored in larger quantities.  
This was also suggested by Power and colleagues [-@powerSEMCalculus2014], who
observed clusters of starches within dental calculus, rather than an even 
distribution across the surface of the dental calculus.

<!--Bacteria also have the ability to adhere to starch [@toppingResistantStarch2003],
which would allow starches to attach to bacterial communities within the biofilm.
Following deposition and/or mineralisation events, the granules that are able to 
resist hydrolysis and 
mechanical removal will then be trapped in pockets within the mineral matrix, and
will temporarily be protected from shearing action and salivary 
clearance. -->
<!-- Unmineralised lacunae and channels within the  
 calculus matrix have been shown to contain viable 
 bacteria, and may also be large enough to contain starch granules. These can 
 range from the width of a single cell, to multiple-cell 
 width [@tanCalculusUltrastructure2004], of which the latter could feasibly contain 
   starch granules.  -->
If this is indeed one of the incorporation mechanisms of starch granules, then the 
limit of incorporated starches is set by the number and size of channels. Since 
this is likely controlled by the size of the calculus deposit, then the number of 
starches will increase as the size of the biofilm increases, which is consistent 
with our results 
(Figures \@ref(fig:cor-plot) and \@ref(fig:cor-plot2)).

>>>>

---
title: "Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model"
author:
  - Bjørn Peare Bartholdy
  - Amanda G. Henry
date: "`r Sys.Date()`"
output:
  bookdown::word_document2:
    reference_docx: "../templates/template.docx"
  bookdown::pdf_document2:
    #template: "../templates/sn-article-template/sn-article.tex"
    includes:
      in_header: "../templates/latex/pre-print/preamble.tex"
      before_body: "../templates/latex/pre-print/before_body.tex"
    df_print: kable
    toc: no
    latex_engine: xelatex
    fig_caption: yes
    number_sections: yes
    keep_tex: true
  bookdown::html_document2:
    df_print: kable
    keep_md: true
linestretch: 2.0 # double-spacing for review purposes
bibliography: refs.bib
csl: apa.csl
abstract: |
 Dental calculus, mineralised plaque, has proven to contain a wealth of information 
 on the dietary habits of past populations. The extraction and identification of
 starch granules contained within dental calculus has already provided valuable
 dietary insights from a wide range of regions and time periods, and seem to preserve 
 well in the archaeological record.
 While the majority of studies have focused on microfossil extraction and
 identification from archaeological remains, few studies have 
 addressed the biases of starch retention in dental calculus, and have been
 limited to *in vivo* studies on modern humans and non-human primates.
 Here, we present a multispecies oral biofilm model, which allows experimental
 research to be conducted on *in vitro* dental calculus in a controlled laboratory
 setting. The biofilms were exposed treatment solutions with known quantities of
 dietary starches (wheat and potato) during the 25-day growth period. After this,
 the starch granules were extracted from the mature biofilm (dissolution in EDTA),
 and counted. We 
 show that the granule counts extracted from the model dental calculus represented 
 a low proportion of the total counts (ranging from 0.064% to 0.161%) exposed to 
 the biofilms throughout the experiment. Additionally, we found that the ratios
 of granule sizes from the extracted starch granules differed from the original
 treatment solutions, with large granules (>20 $\mu$m) consistently being
 under-represented.
 Additionally, the size ratios of granules within starch species shifted between 
 the input  starch solutions and the granules extracted from the model calculus, 
 with large granules  (>20 $\mu$m) consistently being underrepresented. 
 We also found a correlation
 between the absolute granule counts and dry-weight of the biofilm 
 (*r* = 0.66, 90%CI[0.46,0.79]), as well
 as between the concentration (count per mg) of granules and dry-weight 
 (*r* = 0.3, 90%CI[0.06,0.51]).  
 Our results reinforce previous *in vivo* studies suggesting that dental calculus 
 presents a very small, and partly biased picture of the original dietary intake 
 of starches, with an over-representation of plants producing granules smaller than
 20 $\mu$ in size. The experimental model presented here is well-suited to address
 the need for further validation of methods and biases associated with dietary
 research on dental calculus.
keywords: 
  - oral biofilm
  - starch
  - alpha-amylase
  - dental calculus
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.path = "../figures/")
library(tidyverse)
library(broom)
library(here)
library(patchwork)
# load R scripts
source(here("analysis/scripts/analysis.R"))
source(here("analysis/scripts/tables.R"))
source(here("analysis/scripts/figures.R"))
source(here("analysis/scripts/amylase-assay.R"))
## Compile refs.bib using rbbt package ##
cite_keys <- rbbt::bbt_detect_citations(here("analysis/paper/paper.Rmd")) 
#cite_keys <- unlist(cite_keys, use.names = FALSE)
rbbt::bbt_write_bib(here("analysis/paper/refs.bib"), keys = cite_keys, overwrite = T)
## end refs.bib ##

## Word count ##
wd_count <- wordcountaddin::word_count()
```

Word count: `r wd_count`

@@

<!-- more detailed background, comprehensible to scientists in related disciplines -->
Until recently, only a few studies directly investigated the presence of plant
microremains
in the dental calculus of archaeological remains. The ability to extract phytoliths
from the dental calculus of archaeological fauna to investigate diet was first 
noted by Armitage [-@armitageExtractionIdentificationOpal1975], 
and later by Middleton and Rovner [-@middletonExtractionOpalPhytoliths1994],
and Fox and colleagues [-@foxPhytolithCalculus1996]. Starches and phytoliths were
extracted from human dental calculus by Cummings and Magennis [-@cummingsMayanCalculus1997].  
In more recent years, the study of dental calculus has increased exponentially, 
and the wealth of information contained within the mineralised matrix has been, 
to a larger extent, acknowledged. The applications of dental calculus span a wide
variety of archaeological research areas, such as oral microbiome 
characterisation (including pathogens) through the analysis of DNA and proteins 
[@adlerSequencingAncientCalcified2013; @warinnerPathogensHost2014],
microbotanical remains [@henryCalculusSyria2008; @hardyStarchGranulesDental2009;
@mickleburghNewInsightsConsumption2012], 
other organic residues and proteins from dietary compounds 
[@buckleyDentalCalculusCooking2014; @hendyProteomicCalculus2018], 
and nicotine-use [@eerkensDentalCalculusSource2018]. Especially the extraction 
of starch granules has become a rich source of dietary
information, as starch granules have proven to preserve well within dental calculus
over a variety of geographical and temporal ranges 
[@pipernoStarchGrains2008; @henryNeanderthalCalculus2014; @taoWheatCalculus2020;
@jovanovicNeolithicCalculus2021].

@@

```{r protocol-fig, fig.cap="Overview of experiment protocol including the plate setup."}
knitr::include_graphics(here("analysis/figures/protocol_overview.png"))
```

@@

An $\alpha$-amylase (EC 3.2.1.1) activity assay was conducted on artificial
saliva samples collected from the plate wells on days 3, 6, 8, 9, 10, 12, and 14. 
Whole saliva samples were collected on days 0, 3, and 5 as positive controls.
Collected samples were stored at 4&deg;C until the assay was conducted on day 18. 
All samples and standard curves were run in triplicates on two separate plates.
Positive control saliva samples were compared against a standard curve containing 
H~2~O, while artificial saliva samples were compared against a standard curve
containing sterile artificial saliva (due to the colour of artificial saliva).
Two photometric readings were conducted for each plate with a 540 nm filter on a 
Multiskan FC Microplate Photometer (Thermo Scientific 51119000). 
The protocol is a slightly modified version of an Enzymatic Assay of $\alpha$-Amylase
(https://www.sigmaaldrich.com/NL/en/technical-documents/protocol/protein-biology/enzyme-activity-assays/enzymatic-assay-of-a-amylase) [@bernfeldAmylase1955], which measures the amount of 
maltose released from starch by $\alpha$-amylase activity. Results are reported
in units (U) per mL enzyme, where 1 U releases 1 mg of maltose.
For a more detailed protocol, see <!-- protocols.io DOI -->

@@

Extraction of starches from the calculus samples was performed by dissolving the
calculus in 0.5 $\tiny{M}$ ethylenediaminetetraacetic acid (EDTA) 
[@trompEDTACalculus2017; @lemoyneCalculusPretreatments2021; @modiCalculusMethodologies2020], 
and vortexing for 3 days until the sample was completely dissolved. 
Twenty $\mu$l of sample was mounted onto a slide with an 18x18 mm coverslip.
When transferring the sample to the slide, the sample was homogenised using 
the pipette to ensure that the counted transects were representative of the 
whole slide. Wheat and mix counts were extrapolated as described above.
Both wheat and potato granules were divided into three size categories: 
small (<10 $\mu$m), medium (10 -- 20 $\mu$m), and large (>20 $\mu$m).

@@

```{r microscope-fig, fig.cap="Microscope image of potato starches on a microscope slide, as well as a bacterial community. Scale bar = 20 $\\mu$m."}
knitr::include_graphics(here("analysis/figures/starches_w_bar.jpg"))
```

All samples yielded sufficient biofilm growth and starch incorporation to be 
included in the analysis (Figure \@ref(fig:microscope-fig)), resulting in a total of 48 biofilm samples (two plates of 24),
45 of which were used for analysis (three samples were set aside for later
 analysis).
Most control samples contained no starch granules, while some contained negligible
quantities (see Supplementary Material).

@@

```{r amylase-activity}
# table of results reported in units amylase per mL enzyme (but let's be honest,
  # ...it doesn't really matter what the unit is. No activity is no activity)
cols <- c("1", "2", "3")  
rows <- c("S1", "S2", "S3", "B1", "B2", "B3", "B4", "B5", "BT1", "BT2", "BT3")
plt1_ph1_result <- rbind(sal1_ph1, bmm1_ph1)
  plt1_ph1_result[plt1_ph1_result < 0] <- 0

amylase_results <- rbind(sal1_ph1, sal1_ph2, sal2_ph1, sal2_ph2, bmm1_ph1,
                         bmm1_ph2, bmm2_ph1, bmm2_ph2)
amylase_results[amylase_results < 0] <- NA
positive_amylase <- na.omit(amylase_results)
```

<!--

```{r amylase-activity-tbl}
rownames(plt1_ph1_result) <- rows
colnames(plt1_ph1_result) <- cols
knitr::kable(signif(plt1_ph1_result, 3), row.names = TRUE, caption = "Amylase activity in U/mL enzyme, where a U is mg maltose released from starch in six minutes at 36 degrees Celsius. Plate 1, photometric reading 1. Negative values converted to 0.")
```
-->

No $\alpha$-amylase activity was detected in any of the artificial
saliva samples from any of the days that were sampled. Only positive controls
contained amylase activity that could be detected in the assay, ranging from
`r paste(signif(range(positive_amylase),3), collapse = " to ")` U/mL enzyme 
(Supplementary Material).
The results are not comparable to other studies presenting $\alpha$-amylase activity
levels in humans, as the unit definition may differ; 
however, they are sufficient to show that there is no activity in the system.

@@

@@

```{r ratio-plots, fig.cap="Proportion of sizes of starch granules from solutions (outer ring) and treatment samples (inner ring) in separated wheat (A) and potato (B) treatments, and mixed wheat (C) and potato (D) treatments."}
#plot_grid(pl_wheat1, pl_potato1, labels = c("A", "B"), label_y = 0.1)
(pl_wheat1 + pl_potato1) / (pl_wheat2 + pl_potato2) + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

<!--
```{r sep-plot, fig.cap="(A) Proportion (%) of sizes of starch granules in the wheat solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and (B) in the potato solution (outer ring) and extracted from the potato-treatment samples (inner ring)."}
#plot_grid(pl_wheat1, pl_potato1, labels = c("A", "B"), label_y = 0.1)
pl_wheat1 + pl_potato1 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

```{r mix-plot, fig.cap="Proportion of sizes of (A) wheat granules in the mixed solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and sizes of (B) potato granules in the solution (outer ring) and extracted from the potato-treatment samples (inner ring)."}
#plot_grid(pl_wheat2, pl_potato2, labels = c("C", "D"), label_y = 0.1)
pl_wheat2 + pl_potato2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```
 -->

@@

<!-- separated and mixed treatments -->
The difference in incorporation between the size categories resulted in a change
in size ratios between the original starch solutions and the extracted samples.
Large potato granules (> 20 $\mu$m) were most affected, with a 
`r paste0(abs(diff_sep$potato[3]), "%")` 
decrease in relative abundance in the potato-only treatment, and a
`r paste0(abs(diff_mix$potato[3]), "%")` 
decrease in mixed treatments. Medium granules increased in relative abundance 
across all samples, while small granules decreased in wheat treatments and 
increased in potato treatments 
(Figure \@ref(fig:ratio-plots)).

@@

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 $\mu$m in 
diameter were underrepresented in the calculus samples compared to the original 
starch solutions, an effect that was consistent across all three treatments, 
while medium granules (10--20 $\mu$m) were often over-represented
(Table \@ref(tab:sample-prop-tbl), 
and 
Figure \@ref(fig:ratio-plots)).
Large potato granules were most affected, and can reach up to 100 $\mu$m
in diameter, whereas wheat generally only reach up to 35 $\mu$m 
[@gismondiStarchGranulesData2019; @haslamDecompositionStarch2004; @seidemannStarchAtlas1966, 174-176].
This may also suggest that granule morphology plays a role. Large wheat granules 
are lenticular and have a larger surface area compared 
to volume, whereas large potato granules are ovoid and have a larger volume
compared to surface area 
[@vandeveldeStarchMorphology2002; @janeAnthologyStarch1994; @reichertStarchBible1913b, 364-365; @seidemannStarchAtlas1966, 174-176]. 
Another potentially important factor 
from our results is the size of the calculus deposit. We found a 
`r strength_cor` `r direct_cor` correlation between biofilm size and retained 
starch granules (Figure \@ref(fig:cor-plot));
a result that contradicts findings from archaeological contexts
[@wesolowskiEvaluatingMicrofossilContent2010; @dudgeonDietGeographyDrinking2014].
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \@ref(fig:cor-plot2));
while the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches. 
Further research is needed to determine why this differs from previous archaeological
findings.
<!--
Given the low proportion of starch granule retention in our dental calculus model,
and the limited difference between treatments, we can conjecture that starch 
availability is not a main factor in starch incorporation. While there was a
proportional incorporation, i.e. more granules in the solution led to more granules
in the calculus, the difference between solutions [a number/percentage] was of a
higher magnitude than the difference between samples <!-- a number/percentage-->

<!-- Scientific findings as extension of previous work -->
Previous research conducted on dental calculus from contemporary humans and non-human
primates suggest a high level of stochasticity involved in the retention of 
starch granules in dental calculus, and that starch granules extracted from dental 
calculus are underrepresented with regard
to actual starch intake, which is consistent with our findings (illustrated by high
standard deviations and low proportional incorporation).
Leonard and colleagues [-@leonardDentalCalculus2015] found individual
calculus samples to be a poor predictor of diet in a population, as many of the 
consumed plants were missing from some individual samples, but were present in others.   
Power and colleagues [-@powerChimpCalculus2015] 
presented similar findings in non-human primates, where phytoliths were more
representative of individual diets than starch granules.
The size bias is also consistent with the findings by Power and colleagues
[-@powerChimpCalculus2015],
who found that plants producing starches 10--20 $\mu$m in
size were over-represented; although, the representation of granules larger than
20 $\mu$m is unclear.

<!-- add potential for morphological differences affecting incorporation 
 Medium potato starches had the highest rate of incorporation 
 All large starches had a consistently low rate of incorporation 
 Starch granule adhesion to bacteria and enzymes? Do they have a preference?
 Binding to select species in a biofilm could explain clustering
 Colonisation of bacteria onto starch @nagaraBacteriaNutrients2017 (although not
 clear if the bacteria in our model are even 'consuming' starch-->

@@

<!-- How are starch granules trapped in dental calculus? Does anyone actually know? 
-->
The mechanism by which starch granules are incorporated into plaque and calculus 
remains largely unknown, and few studies have directly investigated potential
mechanisms. We know that a proportion of the starch granules entering 
the mouth can become trapped in the plaque/calculus, and can be recovered from
archaeological samples of considerable age 
[@henryNeanderthalCalculus2014; @buckleyDentalCalculusCooking2014; @wuDietEarliest2021]. 
Studies have also shown that not all starch granules come from a dietary source. 
Other pathways include cross-contamination from plant interactions in soil, such 
as palm phytoliths adhering to the skin of sweet potatoes 
[@trompDietaryNondietary2015], 
or accidental ingestion not related to food consumption 
[@radiniFoodMultiplePathways2017; @radiniMedievalWomenEarly2019].  
When starch granules enter the mouth, whether through ingestion of food or accidental
intake, they immediately encounter multiple obstacles. It is likely
that the bulk of starch granules are swallowed along with the food, and are 
only briefly present in the oral cavity. Other granules that are broken off
during mastication may be retained in the dentition through attachment to 
tooth surfaces (including plaque and dental calulus) and mucous membranes 
[@kashketFoodRetention1991; @doddsCarbohydrateRetention1988].
Bacteria also have the ability to adhere to starch granules 
[@toppingResistantStarch2003],
which would allow starches to attach to bacterial communities within the biofilm.
These granules are then
susceptible to mechanical removal by the tongue, salivary clearance, and hydrolysis
by $\alpha$-amylase [@kashketFoodParticles1996].
The susceptibility of granules to hydrolysis depends on the crystallinity and size
of the starch granule, as well as the mode of processing. Smaller and pre-processed
(e.g. cooking) starch granules are more susceptible to enzymatic degradation, 
while dehydrated starches will have a reduced susceptibility
[@lingstromStarchyFood1994; @francoStarchDegradation1992; @haslamDecompositionStarch2004; 
@bjorckStarchProcessing1984; @henryCookingStarch2009].
Cracks on the surface, as well as unmineralised islands and channels exist within 
dental calculus and may also be able to contain starch granules 
[@powerSEMCalculus2014; @tanCalculusUltrastructure2004; @charlierSEMCalculus2010].
Starch granules that are trapped in these pockets are (at least to some extent) 
protected from aforementioned clearance mechanisms, especially once
a new layer of plaque has covered the surface of the plaque/calculus. 
The size bias against large granules (>20 $\mu$m) from both wheat and potato 
(Table \@ref(tab:sample-prop-tbl)) may give further credence to
this incorporation pathway, as the smaller starch granules have an advantage over
larger granules, and can be stored in larger quantities.  
This was also suggested by Power and colleagues [-@powerSEMCalculus2014], who
observed clusters of starches within dental calculus, rather than an even 
distribution across the surface of the dental calculus.  
Granules trapped 
in plaque/calculus may still be susceptible to hydrolysis, as $\alpha$-amylase has
the ability to bind to both tooth enamel and bacteria within a biofilm and retain
a portion of its hydrolytic activity
[@scannapiecoSalivaryAmylase1993; @nikitkovaStarchBiofilms2013; @tanBacterialViability2004; @tanCalculusUltrastructure2004].
After the death of an individual, starches within dental calculus are susceptible 
to further degradation by post-depositional processes, depending on burial environment 
(pH, temperature, moisture content, microorganisms)  
[@henryCookingStarch2009; @graneroStarchTaphonomy2020; @haslamDecompositionStarch2004; @francoStarchDegradation1992].
